---
title: "Single-Cell Analysis Tutorial"
author: "Noah Quimpo Balarbar"
date: "`r Sys.Date()`"
Dataset: GSE180678
output: html_document
---
# Introduction
This notebook documents the end-to-end bioinformatics pipeline for analyzing single-cell RNA sequencing (scRNA-seq) data using the Seurat v5 ecosystem. 
The primary goal is to take raw count data from human heart cells and analyze the transcriptomic response to myocardial infarction at the single-cell level.

## Analysis Objectives
* Quality Control(QC): Filter out low-quality cells based on gene counts (nFeature_RNA) and mitochonrial contamination (percent.mt)
* Normalization and Scaling: Correct raw data to ensure gene expression is comparable across all cells
* Dimensionality Reduction: Utilize Principal Component Analysis (PCA) to simplify the high-dimensionality of gene space, and utilize UMAP for 2D visualization
* Clustering and Annotation: ID clusters of cells with similar transcriptional profiles and annotate them using established biomarkers
```{r setup, include = FALSE}
# knitr::opts_chunk$set(fig.width=7, fig.height=5)
knitr::opts_chunk$set(fig.asp = 0.618) # Golden ratio
```


```{r, message = FALSE, warning = FALSE}
library(languageserver)
library(dplyr)
library(Seurat)
library(patchwork)
```

# Data Setup
```{r, warning = FALSE}
# Rename files appropriately:
# barcodes.tsv.gz
# features.tsv.gz
# matrix.mtx.gz
allData <- Read10X(data.dir = "data/GSE180678_RAW")
cardCells <- CreateSeuratObject(counts = allData, project = "CardCellAnalysis")
cardCells
```

# Pre-processing
## Selecting Viable Cells via Mitochondrial Distress
```{r, warning = FALSE}
# Track Mitochondrial Distress
cardCells[["percent.mt"]] <- PercentageFeatureSet(cardCells, patter = "^MT-")

VlnPlot(cardCells, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

1) **nFeature_RNA plot:** Number of unique genes found in each cell

2) **nCount_RNA:** Number of RNA molecules detected

3) **percent.mt:** Cells in Mitochondrial distress, healthy cells clustered around 5%
    
    - 5% chosen as the mitochondrial threshold


```{r, warning = FALSE}
# Visualize feature-feature relationships:
FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

1) **nCount_RNA vs percent.mt**: Most cells are clusted with low mitochondrial distress

2) **nCount_RNA vs nFeature_RNA**: More RNA molecules is positively correlated with more genes
    - Following Current Best Practices in single-cell RNA-seq analysis, 5000 features chosen


```{r, warning = FALSE}
# Subset data to only include cells that meet 5% mitochondrial threshold
cardCells <- subset(cardCells, subset = nFeature_RNA > 200 & nFeature_RNA <5000 & percent.mt < 5)
```

## SCTransform
SCTransform completes all 3: Normalization, Scaling, and Finding Variable Features all in a single command and replaces these steps in the general workflow.
In addition, SCTransform is also able to remove confounding sources of variation.

```{r, warning = FALSE}
cardCells <- SCTransform(cardCells, vars.to.regress = "percent.mt", verbose = FALSE)
```

::: {#refs}
- Simonson B, Chaffin M, Hill MC, Atwa O, Guedira Y, Bhasin H, Hall AW, Hayat S, Baumgart S, Bedi KC Jr, Margulies KB, Klattenhoff CA, Ellinor PT. Single-nucleus RNA sequencing in ischemic cardiomyopathy reveals common transcriptional profile underlying end-stage heart failure. Cell Rep. 2023 Feb 28;42(2):112086. doi: 10.1016/j.celrep.2023.112086. Epub 2023 Feb 14. PMID: 36790929; PMCID: PMC10423750.

Current Best Practices in single-cell RNA-seq analysis: a tutorial
- Luecken MD, Theis FJ. Current best practices in single-cell RNA-seq analysis: a tutorial. Mol Syst Biol. 2019 Jun 19;15(6):e8746. doi: 10.15252/msb.20188746. PMID: 31217225; PMCID: PMC6582955.
:::