---
title: "Single-Cell Analysis Tutorial"
author: "Noah Quimpo Balarbar"
date: "`r Sys.Date()`"
Dataset: GSE180678
output: html_document
---
# Introduction
This notebook documents the end-to-end bioinformatics pipeline for analyzing single-cell RNA sequencing (scRNA-seq) data using the Seurat v5 ecosystem. 
The primary goal is to take raw count data from human heart cells and analyze the transcriptomic response to myocardial infarction at the single-cell level.

## Analysis Objectives
* Quality Control(QC): Filter out low-quality cells based on gene counts (nFeature_RNA) and mitochonrial contamination (percent.mt)
* Normalization and Scaling: Correct raw data to ensure gene expression is comparable across all cells
* Dimensionality Reduction: Utilize Principal Component Analysis (PCA) to simplify the high-dimensionality of gene space, and utilize UMAP for 2D visualization
* Clustering and Annotation: ID clusters of cells with similar transcriptional profiles and annotate them using established biomarkers

```{r setup, include = FALSE}
# knitr::opts_chunk$set(fig.width=7, fig.height=5)
knitr::opts_chunk$set(fig.asp = 0.618) # Golden ratio
```


```{r, message = FALSE, warning = FALSE}
library(languageserver)
library(tidyverse)
library(dplyr)
library(Seurat)
library(patchwork)
install.packages('BiocManager')
```

# Data Setup
```{r, warning = FALSE}
# Rename files appropriately:
# barcodes.tsv.gz
# features.tsv.gz
# matrix.mtx.gz
allData <- Read10X(data.dir = "data/GSE180678_RAW")
cardCells <- CreateSeuratObject(counts = allData, project = "CardCellAnalysis")
cardCells
```

# Pre-processing
## Selecting Viable Cells via Mitochondrial Distress
```{r, warning = FALSE}
# Track Mitochondrial Distress
cardCells[["percent.mt"]] <- PercentageFeatureSet(cardCells, patter = "^MT-")

VlnPlot(cardCells, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

1) **nFeature_RNA plot:** Number of unique genes found in each cell

2) **nCount_RNA:** Number of RNA molecules detected

3) **percent.mt:** Cells in Mitochondrial distress, healthy cells clustered around 5%
    
    - 5% chosen as the mitochondrial threshold


```{r, warning = FALSE}
# Visualize feature-feature relationships:
FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

1) **nCount_RNA vs percent.mt**: Most cells are clusted with low mitochondrial distress

2) **nCount_RNA vs nFeature_RNA**: More RNA molecules is positively correlated with more genes
    - Following Current Best Practices in single-cell RNA-seq analysis, 5000 features chosen

Further QC questions:

1) How many genes are being expressed per cell by quantile?

2) How much mitochondrial stress is being expressed per cell by quantile?

These questions will help determine how to QC viable cells
```{r, warning = FALSE}
# Check quantiles of features
quantile(cardCells$nFeature_RNA, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

quantile(cardCells$percent.mt, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))
```
See here that for this specific dataset, the mean percent.mt is about 8%.

Given that cardiac cells generally have higher percent.mt due to features such as higher mitochondrial abundance, 
in conjunction with the cells originating from a patient with ischemic cardiomyopathy, a higher percent.mt cutoff is chosen (15%)


```{r, warning = FALSE}
# Subset data to only include cells that meet 5% mitochondrial threshold
dim(cardCells)
cardCells <- subset(cardCells, subset = nFeature_RNA > 50 & nFeature_RNA <3500 & percent.mt < 15)
dim(cardCells)
```

## SCTransform
SCTransform completes all 3: Normalization, Scaling, and Finding Variable Features all in a single command and replaces these steps in the general workflow.
In addition, SCTransform is also able to remove confounding sources of variation.

NOTE: Install package "BiocManager"

```{r, warning = FALSE}
cardCells <- SCTransform(cardCells, vars.to.regress = "percent.mt", verbose = FALSE)
```

## Principal Component Analysis (PCA)
```{r, warning = FALSE}
cardCells <- RunPCA(cardCells, features = VariableFeatures(object = cardCells))
```

Lets view which genes taht are expressed (positively and negatively) on the end of each PC axis

```{r, warning = FALSE}
print(cardCells[["pca"]], dims = 1:5, nfeatures = 5)
```

Then, lets look at the loading score, or which genes are contributing the most to the first 2 PC's
```{r, warning = FALSE}
VizDimLoadings(cardCells, dims = 1:2, reduction = "pca")
```

Now lets look at clusters of cells on the first 2 PC's

```{r, warning = FALSE}
DimPlot(cardCells, reduction = "pca") + NoLegend()
```

Use an ElbowPlot to determine the number of PC's needed for clustering
```{r, warning = FALSE}
ElbowPlot(cardCells)
```

17 PC's will be utilized

Now cluster the cells over the 17 PC's
```{r, warning = FALSE}
cardCells <- FindNeighbors(cardCells, dims = 1:17)
cardCells <- FindClusters(cardCells, resolution = 0.5) 

head(Idents(cardCells), 5)
```

# UMAP
```{r, warning = FALSE}
cardCells <- RunUMAP(cardCells, dims = 1:10)
DimPlot(cardCells, reduction = "umap", , pt.size = 2.5)
```

See genes are being expressed by each cluster
```{r, warning = FALSE}
clusterMarkers <- FindAllMarkers(cardCells, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

clusterMarkers %>% group_by(cluster) %>% top_n(n = 2, wt = avg_log2FC)
```

Cross-reference the top two genes expressed by each cluster with the Panglaodb

```{r, warning = FALSE}

```

::: {#refs}
- Simonson B, Chaffin M, Hill MC, Atwa O, Guedira Y, Bhasin H, Hall AW, Hayat S, Baumgart S, Bedi KC Jr, Margulies KB, Klattenhoff CA, Ellinor PT. Single-nucleus RNA sequencing in ischemic cardiomyopathy reveals common transcriptional profile underlying end-stage heart failure. Cell Rep. 2023 Feb 28;42(2):112086. doi: 10.1016/j.celrep.2023.112086. Epub 2023 Feb 14. PMID: 36790929; PMCID: PMC10423750.

Current Best Practices in single-cell RNA-seq analysis: a tutorial
- Luecken MD, Theis FJ. Current best practices in single-cell RNA-seq analysis: a tutorial. Mol Syst Biol. 2019 Jun 19;15(6):e8746. doi: 10.15252/msb.20188746. PMID: 31217225; PMCID: PMC6582955.

Single Cell RNA sequencing Database (cross reference genes)
- https://panglaodb.se/
:::