---
title: "Single-Cell Analysis Tutorial"
author: "Noah Quimpo Balarbar"
date: "`r Sys.Date()`"
Dataset: GSE180678
output: html_document
---
```{r}
library(languageserver)
library(dplyr)
library(Seurat)
library(patchwork)
```
# Introduction
This notebook documents the end-to-end bioinformatics pipeline for analyzing single-cell RNA sequencing (scRNA-seq) data using the Seurat v5 ecosystem. 
The primary goal is to take raw count data from human heart cells and analyze the transcriptomic response to myocardial infarction at the single-cell level.

## Analysis Objectives
* Quality Control(QC): Filter out low-quality cells based on gene counts (nFeature_RNA) and mitochonrial contamination (percent.mt)
* Normalization and Scaling: Correct raw data to ensure gene expression is comparable across all cells
* Dimensionality Reduction: Utilize Principal Component Analysis (PCA) to simplify the high-dimensionality of gene space, and utilize UMAP for 2D visualization
* Clustering and Annotation: ID clusters of cells with similar transcriptional profiles and annotate them using established biomarkers


# Data Setup
```{r}
# Rename files appropriately:
# barcodes.tsv.gz
# features.tsv.gz
# matrix.mtx.gz
allData <- Read10X(data.dir = "data/GSE180678_RAW")
cardCells <- CreateSeuratObject(counts = allData, project = "CardCellAnalysis")
cardCells
```

# Pre-processing
## Selecting Viable Cells via Mitochondrial Distress
```{r}
# Track Mitochondrial Distress
cardCells[["percent.mt"]] <- PercentageFeatureSet(cardCells, patter = "^MT-")

# Keep cells with <5% mitochondrial distress
plot1 <- VlnPlot(cardCells, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", ncol = 3))

# Visualize feature-feature relationships:
plot2 <- FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot3 <- FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

plot2 + plot3
```

```{r}
# Visualize feature-feature relationships:
plot1 <- FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(cardCells, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```



::: {#refs}
- Simonson B, Chaffin M, Hill MC, Atwa O, Guedira Y, Bhasin H, Hall AW, Hayat S, Baumgart S, Bedi KC Jr, Margulies KB, Klattenhoff CA, Ellinor PT. Single-nucleus RNA sequencing in ischemic cardiomyopathy reveals common transcriptional profile underlying end-stage heart failure. Cell Rep. 2023 Feb 28;42(2):112086. doi: 10.1016/j.celrep.2023.112086. Epub 2023 Feb 14. PMID: 36790929; PMCID: PMC10423750.
:::